## Project 1

### Introduction

This data is from a personal activity monitoring device which collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data is downloaded from the course web site:
Dataset: Activity monitoring data

```{r setoptions}
knitr::opts_chunk$set(fig.width=5, fig.height=3) 
```

```{r}
library(lattice)
library(dplyr)
```

```{r loading}
data <- read.csv("activity.csv")
```
The dataset contains 17568 observations and 3 variables:
        **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA);
        **date**: The date on which the measurement was taken in YYYY-MM-DD format;
        **interval**: Identifier for the 5-minute interval in which measurement was taken

### The total number of steps per day

I summed the number of steps per day (61 days). Here is the histogram of the total number of steps taken each day.
```{r sum_steps}
steps_day <- tapply(data$steps, data$date, sum, na.rm=TRUE)
histogram(steps_day, nint=15, type="count", xlab="Total steps per day")
```

```{r mean of steps per day}
mean_st <- mean(steps_day)
```

The mean of the total number of steps taken per day is `r mean_st` .

```{r median of steps per day}
median_st <- median(steps_day)
```

And the median of the total number of steps taken per day is `r median_st` .

### Average daily activity pattern

I calculated the average number of steps taken in each interval across all days created a plot of the 5-minute interval and the average number of steps taken, averaged across all days.

```{r}
data$interval <- as.factor(data$interval)
steps_int <- tapply(data$steps, data$interval, mean, na.rm=TRUE)
lvl <- as.integer(levels(data$interval))
df <- data.frame(lvl, steps_int)
df <- arrange(df,lvl)
xyplot(steps_int~lvl,df, type="l", xlab="Interval", ylab="Average number of steps")
```

```{r interval with max mean of steps}
maxsteps_int <- df[1][df$steps_int %in% c(max(steps_int)),]
```

The interval containing the maximum number of steps is `r maxsteps_int` .

### Imputing missing values

```{r number of NAs}
total_na <- sum(is.na(data$steps))
```

Total number of missing values is `r total_na` .

I decided to replace the missing values in variable "steps" with the mean for that 5-minute interval. In order to do it I created a new variable "gr_mean" containing the means values for the intervals. The new vector "full_steps" contains the values of "steps" with replaced NA and data frame "data_full" is "data" with "full_steps" added to it.

```{r filling NA}
data_g <- data %>% group_by(interval) %>% mutate(gr_mean=mean(steps, na.rm=TRUE))
full_steps <- ifelse(!is.na(data_g$steps), data_g$steps, data_g$gr_mean)
data_full <- data.frame(full_steps, data)
data_full <- select(data_full, -c(steps))
```

The new data frame with filled values "data_full" looks like this:

```{r}
head(data_full)
```

And here is the histogram of total number of steps taken per day with the missing data filled in.

```{r histogram full_steps}
fullst_day <- tapply(data_full$full_steps, data_full$date, sum)
histogram(fullst_day, nint=15, type="count", xlab="Total steps per day with filled NA")
```

```{r mean of steps per day (full)}
mean_fullst <- mean(fullst_day)
```

The mean of the total number of steps taken per day is `r mean_fullst` .

```{r median of steps per day (full)}
median_fullst <- mean(fullst_day)
```

The median of the total number of steps taken per day is `r median_fullst` .

The values from data with filled NA's differ from the values from data with NA's. We can see this in the table here.

```{r comparasion}
tab <- data.frame("Mean"=c(mean_st, mean_fullst), "Median"=c(median_st, median_fullst), row.names = c("Steps", "Full steps") )
tab
```

### Differences in activity patterns between weekdays and weekends

I replaced the dates in variable "date" with the days of the week and created a new factor variable in the dataset with levels "weekday" and "weekend".

```{r new factor weekday/weekend}
w_days <- weekdays(as.Date(data_full$date))
days <- as.factor(ifelse(w_days=="Saturday"|w_days=="Sunday", "weekend", "weekday"))
data_full <- cbind(data_full, days)
head(data_full)
```

Here is a plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days. 

```{r plot_weekend/weekday}
data_full <- data_full %>% group_by(interval) %>% mutate(full_mean=mean(full_steps))
xyplot(data_full$full_mean~data_full$interval | data_full$days, layout=c(1,2), type="l", xlab="Interval", ylab="Number of steps")
```

